---
params:
  library: Library
  processed_sce: "../results/scpca_processed_data/SCPCS000216/SCPCL000290_processed_sce_clustering.rds"
  stats_dir: "../results/scpca_processed_data/SCPCS000216/clustering_stats"
  cluster_type: "louvain"
  nearest_neighbors_range: 5:25
  nearest_neighbors_increment: 5
  project_root: NULL
  date: !r Sys.Date()

title: "`r glue::glue('Clustering analysis report for {params$library}')`"
author: "CCDL"
date: "`r params$date`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: hide
---

This notebook will look at additional clustering parameters on the expression of
the genes in a single sample of the dataset of interest, from an unbiased approach.

## Set Up

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

# If project root is not provided use here::here() 
if (is.null(params$project_root)) {
  project_root <- here::here()
} else {
  project_root <- params$project_root
} 

# Source in set up function
source(file.path(project_root, "utils", "setup-functions.R"))
source(file.path(project_root, "utils", "clustering-functions.R"))

# Load project
setup_renv(project_filepath = project_root)

# Set theme for plotting
library(ggplot2)
theme_set(theme_bw())
```

```{r}
# Load libraries
library(readr)
library(SingleCellExperiment)
library(dplyr)
library(miQC)
library(ggpubr)
library(scater)
```

```{r}
# Read in SCE object
processed_sce <- read_rds(file.path(params$processed_sce))
```

## Plots

## UMAP plot results for `r params$library`

Below is a UMAP plot where each cell has been colored based on the cell's cluster assignment. 
Here clustering was performed using `r params$cluster_type` clustering.

```{r}
# If param$nearest_neighbors_range is a single value, set nearest_neighbors_increment to NULL
if(length(params$nearest_neighbors_range) == 1) {
  nearest_neighbors_increment <- 1
}
# If a nearest neighbors increment exists, then create a sequence for clustering 
if(!(params$nearest_neighbors_increment == 1)){
  nn_range <- seq(min(params$nearest_neighbors_range), 
                  max(params$nearest_neighbors_range), 
                  params$nearest_neighbors_increment) 
  # If no nearest neighbors increment has been input then the provided range is 
  # directly used for clustering 
} else {
  nn_range <- params$nearest_neighbors_range
}

# Grab column name with clustering results
cluster_column_names <- paste(params$cluster_type, nn_range, sep = "_")
existing_columns <- intersect(cluster_column_names, colnames(colData(processed_sce)))

# Check length of `existing_columns`
if(length(existing_columns) == 0){
  stop("There are no existing clustering results, therefore a report cannot be generated.")
}

# Set number of columns for plotting and use this to define figure heights
num_col <- 2
figure_height <- max(ceiling(length(existing_columns)/num_col) * 5, 4)

knitr::opts_chunk$set(
  fig.height = figure_height
)
```

```{r}
# Plot UMAP coordinatates
plot_list <- existing_columns %>%
  purrr::map(
    ~ plotReducedDim(processed_sce, dimred = "UMAP", colour_by = .x) +
      theme_bw() +
      theme(text = element_text(size = 8))
  )

cowplot::plot_grid(plotlist = plot_list, ncol = num_col)
```

## Plot cluster validity stats

```{r}
# Grab the cluster validity stats data frame
all_validity_stats_df <- readr::read_tsv(file.path(
  params$stats_dir,
  paste0(params$library, "_clustering_all_validity_stats.tsv")
))
```

#### Purity plots

```{r}
# Plot individual cluster purity stats
purity_plots <- plot_cluster_purity(all_validity_stats_df, num_col)

purity_plots + theme(text = element_text(size = 8))
```

#### Silhouette width plots

```{r}
# Plot individual cluster silhouette width stats
silhouette_plots <- plot_cluster_silhouette_width(all_validity_stats_df, num_col)

silhouette_plots + theme(text = element_text(size = 8))
```

### Summary plots

```{r}
knitr::opts_chunk$set(
  fig.height = 6
)
```

```{r}
# Grab the cluster validity stats summary data frame
summary_validity_stats_df <- readr::read_tsv(file.path(
  params$stats_dir,
  paste0(params$library, "_clustering_summary_validity_stats.tsv")
))

# purity summary plot
plot_avg_validity_stats(summary_validity_stats_df, "avg_purity")

#silhouette width summary plot
plot_avg_validity_stats(summary_validity_stats_df, "avg_width")
```

## Plot cluster stability

```{r}
# Grab the cluster stability stats summary data frame
summary_stability_stats_df <- readr::read_tsv(file.path(
  params$stats_dir,
  paste0(params$library, "_clustering_summary_stability_stats.tsv")
))

# plot cluster stability ARI values
plot_cluster_stability_ari(summary_stability_stats_df)
```

## Session info
<details>
<summary>R session information</summary>
```{r}
sessionInfo()
```
