---
params:
  library: "library01"
  normalized_sce: "example-results/sample01/library01_processed_sce.rds"
  date: !r Sys.Date()
  goi_input_directory: "example-results/sample01/goi-stats"
  project_root: NULL

title: "`r glue::glue('Gene expression plots for {params$library}')`"
author: "CCDL"
date: "`r params$date`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: false
    code_folding: hide
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

# If project root is not provided use here::here() 
if (is.null(params$project_root)) {
  project_root <- here::here()
} else {
  project_root <- params$project_root
} 

# Tell knitr to knit from the project root to be able to load `renv` properly
# This path goes into effect for SUBSEQUENT chunks, hence the need for 2 chunks
knitr::opts_knit$set(root.dir = project_root)
```

```{r}
# Load libraries
library(magrittr)
library(readr)
library(SingleCellExperiment)
library(scater)
library(ggplot2)
library(ComplexHeatmap)

# Set seed for heatmap colors
set.seed(201)

# Source custom functions script
source(file.path(project_root, "utils", "provided-goi-analysis-functions.R"))
```

```{r}
normalized_sce <- read_rds(file.path(params$normalized_sce))

goi_list <-
  read_tsv(file.path(
    project_root,
    params$goi_input_directory,
    paste0(params$library, "_mapped_genes.tsv")
  ))

goi_df <- as.data.frame(goi_list)
```

```{r}
heatmap_matrix <- Matrix::readMM(file.path(
            params$goi_input_directory,
            paste0(params$library, "_normalized_zscores.mtx")
          ))
# Convert to standard matrix for heatmap plotting
heatmap_matrix <- as.matrix(heatmap_matrix)

column_annotation <- read_rds(file.path(
  project_root,
  params$goi_input_directory,
  paste0(params$library, "_heatmap_annotation.rds")
))
```

```{r results='asis'}
existing_goi <- goi_df %>%
  dplyr::filter(sce_rownames_identifier %in% rownames(logcounts(normalized_sce)) | 
                  gene_id %in% rownames(logcounts(normalized_sce))) %>%
  dplyr::pull(gene_id)

existing_goi_print <- paste(existing_goi, collapse = ", ")
glue::glue("
    <div class=\"alert alert-warning\">
    
    The following genes of interest can be found in the provided dataset:
    
    {existing_goi_print}
    
    </div>
  ")
```

### Heatmap plotting

The below heatmap displays genes on the x-axis, cells on the y-axis, and color indicates gene expression scaled using a z-score. 
Using only the provided genes of interest, hierarchical clustering of cells was performed to identify if any clear structure or groupings of cells with similar gene expression patterns are present. 
The colored bar on the top of the graph indicates which gene set the provided gene of interest corresponds to. 

```{r, fig.width=10, fig.height=10}
# create the ComplexHeatmap
Heatmap(heatmap_matrix,
        # implement a correlation based distance metric
        clustering_distance_rows = "pearson",
        # calculate the average distance between each cluster before merging
        clustering_method_rows = "average",
        clustering_distance_columns = "pearson",
        clustering_method_columns = "average",
        show_row_names = FALSE,
        show_column_names = TRUE,
        top_annotation = column_annotation,
        name = "z score")
```

## Single gene expression plots

### Sina plots

The plot below displays the normalized and transformed expression for each of the provided
genes of interest in comparison to the mean expression of all genes present in the dataset. 

```{r}
if(all.equal(goi_list$gene_id, goi_list$sce_rownames_identifier) == TRUE) {
  gene_id_column <- NULL
} else {
  gene_id_column <- "gene_id"
}

plot_goi_expression_sina(normalized_sce,
                         goi_df,
                         "sce_rownames_identifier",
                         gene_id_column)
```

### UMAP plots

Each plot below represents the UMAP results calculated using this sample's data.
Each dot represents a cell and the color indicates the individual gene of interest's expression (each plot being relevant to a single gene of interest).

```{r}
plot_goi_expression_umap(normalized_sce,
                         goi_df,
                         "sce_rownames_identifier",
                         gene_id_column)
```

### PCA plots

Each plot below represents the PCA results calculated using this sample's data.
Each dot represents a cell and the color indicates the individual gene of interest's expression (each plot being relevant to a single gene of interest).

```{r}
plot_goi_expression_pca(normalized_sce,
                        goi_df,
                        "sce_rownames_identifier",
                        gene_id_column)
```

## Session Info

```{r session_info}
sessionInfo()
```

