---
params:
  library: Library
  pre_processed_sce: "data/Gawad_processed_data/SCPCS000216/SCPCL000290_filtered.rds"
  processed_sce: "results/Gawad_processed_data/SCPCS000216/SCPCL000290_processed.rds"
  cluster_type: "louvain"
  nearest_neighbors: 10
  project_root: NULL
  mito_file: "reference-files/Homo_sapiens.GRCh38.104.mitogenes.txt"
  date: !r Sys.Date()

title: "`r glue::glue('Core analysis report for {params$library}')`"
author: "CCDL"
date: "`r params$date`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: false
    code_folding: hide
---

This notebook will plot the filtering and clustering results from the core analysis workflow.

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

# If project root is not provided use here::here() 
if (is.null(params$project_root)) {
  project_root <- here::here()
} else {
  project_root <- params$project_root
} 

# Tell knitr to knit from the project root to be able to load `renv` properly
# This path goes into effect for SUBSEQUENT chunks, hence the need for 2 chunks
knitr::opts_knit$set(root.dir = project_root)

# Source in set up function
source(file.path(project_root, "utils", "setup-functions.R"))
# Source filtering functions 
source(file.path(project_root, "utils", "filtering-functions.R"))
```

```{r include=FALSE}
# Load project
setup_renv(project_filepath = project_root)

# Set theme for plotting
library(ggplot2)
theme_set(theme_bw())
```

```{r include=FALSE}
# Load libraries
library(readr)
library(SingleCellExperiment)
library(dplyr)
library(miQC)
library(ggpubr)
library(scater)
```

```{r include=FALSE}
# Read in SCE objects
pre_processed_sce <- read_rds(file.path(params$pre_processed_sce))

processed_sce <- read_rds(file.path(params$processed_sce))
```

## Filtering results for `r params$library`

```{r}
# Extract sce metadata
processed_meta <- metadata(processed_sce)

# Define number of cells
num_cells_pre_processed <- dim(pre_processed_sce)[2]
num_filtered_cells <- processed_meta$num_filtered_cells_retained

# Define sample id
if(is.null(processed_meta$sample)){
  sample <- NA
} else {
  sample <- processed_meta$sample
}

# Put a sample information data frame together
sample_information <- tibble::tibble(
  "Sample id" = sample,
  "Library id" = params$library,
  "Cells before filtering" = 
    format(num_cells_pre_processed, big.mark = ',', scientific = FALSE),
  "Cells after filtering" = 
    format(num_filtered_cells, big.mark = ',', scientific = FALSE),
  "Filtering type" =
    format(processed_meta$scpca_filter_method, big.mark = ',', scientific = FALSE),
  "Minimum genes cutoff" = 
        format(processed_meta$min_gene_cutoff, big.mark = ',', scientific = FALSE)
)

if (processed_meta$scpca_filter_method == "manual"){

  sample_information <- sample_information %>%
    mutate(
      "UMI count cutoff" = 
        format(processed_meta$umi_count_cutoff, big.mark = ',', scientific = FALSE),
      "Percent mito cutoff" = 
        format(processed_meta$mito_percent_cutoff, big.mark = ',', scientific = FALSE)
    )
}

if (processed_meta$scpca_filter_method == "miQC" ){

  sample_information <- sample_information %>%
    mutate("Probability compromised cutoff" = 
             format(processed_meta$probability_compromised_cutoff, big.mark = ',', scientific = FALSE))
}

sample_information <- sample_information %>%
  mutate(across(.fns = ~ifelse(.x == "NULL", "N/A", .x))) %>% # reformat nulls
  t()

# Make the table with sample information
knitr::kable(sample_information, align = 'r') %>%
  kableExtra::kable_styling(bootstrap_options = "striped",
                            full_width = FALSE,
                            position = "left") %>%
  kableExtra::column_spec(2, monospace = TRUE)
```

### Filtering plots

The filtering plots below represent the cutoffs used to filter and remove low-quality cells.
When `miQC` filtering is implemented, two plots are displayed.
`miQC` relies on fitting a mixture model using the number of genes expressed by a cell and the percentage of mitochondrial reads. 
The first plot will show the model fit by `miQC` with two model fit lines and each cell being colored by the probability of that cell being compromised.
The second plot indicates which cells have been removed based on the cutoff used for a cell's probability of being compromised.
Cells labeled with `TRUE` are kept for downstream analysis, while all other cells were removed.
When `manual` filtering is implemented, one plot is displayed.
The plot in this case represents the manual filtering cutoffs for the minimum number of genes detected per cell and the minimum unique molecular identifiers (UMI) per cell.
The lines on the plot indicate the cutoffs, where cells with number of genes expressed and unique molecular identifiers (UMI) that are both above the minimum cutoffs are kept for downstream analysis.
Additionally, cells are colored by percent mitochondrial reads. 
Cells with percent mitochondrial reads above the provided cutoff are removed prior to downstream analysis.

Here, cells were filtered by `r processed_meta$scpca_filter_method`.


```{r}
# Read in mito genes
mito_genes <- unique(readLines(params$mito_file))

if (is.null(pre_processed_sce$detected)) {
  # Add cell QC metrics to pre processed sce needed for plotting
  pre_processed_sce <- scpcaTools::add_cell_mito_qc(pre_processed_sce, mito_genes)
}
```


```{r}
if (processed_meta$scpca_filter_method == "manual"){
  # Plot manual filtering thresholds
  filtered_cell_plot <- plot_manual_filtering(
    sce = pre_processed_sce,
    detected_gene_cutoff = processed_meta$detected_gene_cutoff,
    umi_count_cutoff = processed_meta$umi_count_cutoff
  )
}
if (processed_meta$scpca_filter_method == "miQC" ){
  # Remove `prob_compromised` if it exists
  pre_processed_sce$prob_compromised <- NULL
  # Plot model
  filtered_model_plot <- miQC::plotModel(pre_processed_sce, processed_meta$miQC_model)
  # Plot filtering
  filtered_cell_plot <- miQC::plotFiltering(pre_processed_sce, 
                                            processed_meta$miQC_model,
                                            posterior_cutoff = processed_meta$probability_compromised_cutoff)
  # Combine plots
  filtered_cell_plot <-
    ggarrange(filtered_model_plot,
              filtered_cell_plot,
              ncol = 1,
              nrow = 2)
}
  
filtered_cell_plot
```

## Clustering results for `r params$library`

Below is a UMAP plot where each cell has been colored based on the cell's cluster assignment. 
Here clustering was performed using `r params$cluster_type` clustering with a nearest neighbors value of `r params$nearest_neighbors`.

```{r}
# Grab column name with clustering results
cluster_name <- paste(params$cluster_type, params$nearest_neighbors, sep = "_")

# Plot
plotReducedDim(
  processed_sce,
  dimred = "UMAP",
  colour_by = cluster_name,
  text_by = cluster_name,
  text_size = 4,
  point_size = 0.4,
  point_alpha = 0.5
) +
  theme_bw() +
  labs(
    caption = paste0(
      stringr::str_to_title(params$cluster_type),
      " clustering with nearest neighbors value ",
      params$nearest_neighbors
    )
  ) +
  guides(col = guide_legend("Cluster assignment", override.aes = list(size = 3))) +
  theme(text = element_text(size = 14),
        plot.caption = element_text(hjust = 0.5))
```


## Session info
<details>
<summary>R session information</summary>
```{r}
sessionInfo()
```
